---
title: "HW1_text_as_data"
author: "Ishita Vaid"
date: "2026-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#call packages

library("readr")
library("dplyr")
library("tidyr")
library("stringr")
library("tidytext")
library("ggplot2")
library("forcats")
library("tibble")
library("scales")
```


```{r}
#  file paths to call files
file_a <- "texts/A07594__Circle_of_Commerce.txt"
file_b <- "texts/B14801_Free_Trade.txt"

# Read the raw text files into R
text_a <- read_file(file_a)
text_b <- read_file(file_b)

# Combine into a tibble for tidytext workflows
texts <- tibble(
  doc_title = c("Text A", "Text B"),
  text = c(text_a, text_b) 
)

texts
```

```{r}
#number of characters in each document, word tokens, word types after lowercasing (before stopword removal)

corpus_diagnostics <- texts %>%
mutate(n_chars = str_length(text)) %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_to_lower(word)) %>%
  group_by(doc_title) %>%
  summarise(
    n_chars = first(n_chars),
    n_word_tokens = n(),
    n_word_types = n_distinct(word),
    .groups = "drop"
  )

corpus_diagnostics

```
#Text A and Text B are not comparable in length. Text A contains 33,889 word tokens and 191,605 characters while Text B contains 19,922 word tokens/words and 114,211 characters. This means Text A is much longer than Text B in terms of total words and raw word frequency counts will be higher in Text A simply due to its greater length. Therefore, direct comparisons of raw frequencies could deliver misleading results. Due to this, it would be better methodologically to compare relative word frequencies of the documents instead.


```{r}
# stop words removal
data("stop_words")

# Add project-specific stopwords
custom_stopwords <- tibble(
  word = c(
    "vnto", "haue", "doo", "hath", "bee", "ye", "thee"
  )
)

all_stopwords <- bind_rows(stop_words, custom_stopwords) %>%
  distinct(word)

all_stopwords %>% slice(1:10)

```

```{r}
#word counts

word_counts <- texts %>%
  unnest_tokens(word, text) %>%
  mutate(word = str_to_lower(word)) %>%
  anti_join(all_stopwords, by = "word") %>%
  count(doc_title, word, sort = TRUE)

word_counts
```

```{r}
#total number of words (after removing stopwords)

doc_lengths <- word_counts %>%
  group_by(doc_title) %>%
  summarise(
    total_words = sum(n),
    .groups = "drop"
  )

doc_lengths
```

```{r}
#relative frequency of words in documents

word_counts_normalized <- word_counts %>%
  left_join(doc_lengths, by = "doc_title") %>%
  mutate(relative_freq = n / total_words)

word_counts_normalized
```

```{r}
#filter for trade in both texts

trade_freq <- word_counts_normalized %>%
  filter(word == "trade")

trade_freq
```

```{r}
#raw frequency plot (week 2)

plot_n_words <- 20  # you can change this as needed

# Select the most frequent words overall
word_comparison_tbl <- word_counts %>%
  pivot_wider(
    names_from = doc_title,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(max_n = pmax(`Text A`, `Text B`)) %>%
  arrange(desc(max_n))

word_plot_data <- word_comparison_tbl %>%
  slice_head(n = plot_n_words) %>%
  pivot_longer(
    cols = c(`Text A`, `Text B`),
    names_to = "doc_title",
    values_to = "n"
  ) %>%
  mutate(word = fct_reorder(word, n, .fun = max))

ggplot(word_plot_data, aes(x = n, y = word)) + #black magic happens thanks to ggplot
  geom_col() +
  facet_wrap(~ doc_title, scales = "free_x") +
  labs(
    title = "Most frequent words (stopwords removed)",
    subtitle = paste0(
      "Top ", plot_n_words,
      " words by maximum frequency across both texts"
    ),
    x = "Word frequency",
    y = NULL
  ) +
  theme_minimal()

```


```{r}
#relative frequency plot

plot_n_words <- 20  

# Identify top 20 words using raw counts 
top_words <- word_counts %>%
  pivot_wider(
    names_from = doc_title,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(max_n = pmax(`Text A`, `Text B`)) %>%
  arrange(desc(max_n)) %>%
  slice_head(n = plot_n_words) %>%
  select(word)


word_plot_data_norm <- word_counts_normalized %>%
  semi_join(top_words, by = "word") %>%
  mutate(word = fct_reorder(word, relative_freq, .fun = max))


ggplot(word_plot_data_norm, aes(x = relative_freq, y = word)) +
  geom_col() +
  facet_wrap(~ doc_title, scales = "free_x") +
  labs(
    title = "Most frequent words (relative to document length )",
    subtitle = paste0(
      "Top ", plot_n_words,
      " words by maximum raw frequency across both texts"
    ),
    x = "Relative frequency",
    y = NULL
  ) +
  theme_minimal()

```

#Answer 1: Although Text A uses the word trade more times in raw counts (232 vs. 185), Text B uses it more proportionally. The relative frequency of “trade” in Text B (0.0231) is greater than the relative frequency of “trade” in Text A (0.0158). This means that even though Text A mentions “trade” more times cumulatively, trade consists of a larger share of Text B’s total vocabulary. The raw counts suggested that Text A emphasized “trade” more, but normalization reveals that Text B actually emphasizes it more relative to its size.

#Answer 2: If we normalized by the original document length (before removing stopwords), the relative frequencies would be much smaller because the denominator would include many high-frequency function words like “and” and “the”. This would make the proportions less reflective of the texts’ substantive content, since stopwords generally do not contribute to a text’s meaning. As a result, the normalized values would be less representative of the importance/prominence of the word “trade” within the meaningful vocabulary of the document.

